### 2018-07-30 chapters/01-introduction/intro.tex
Implemented all suggestions by Howard

Even with embedded semantic information, a translation between the systems can be difficult. A typical example of complex problems occurs for multivalued functions~\parencite{AISC:MultivaluedFunctions}. A \gls*{cas} usually defines \textit{branch cuts} to compute principal values of multivalued functions~\parencite{Maple:Cuts}, which makes the implementation of a theoretically continues function to a discontinued presentation of it. In general, positioning branch cuts follows conventions, but can be positioned arbitrarily in many cases. Communicate and explain the decision of defined branch cuts is a critical point for \gls*{cas} and can vary between the systems~\parencite{Branches:acot}. Figure~\ref{fig:acot-cut-compare} illustrates two examples of different branch cut positioning for the inverse trigonometric arccotangent function. While \Maple{} defines the branch cut at ${[-\infty\iunit, -\iunit]}$, ${[\iunit,\infty\iunit]}$ (figure~\ref{fig:acot-cut1}), \Mathematica{} defines the branch cut at ${[-\iunit, \iunit]}$ (figure~\ref{fig:acot-cut2}).

\begin{figure}[ht]
    \centering
    \subfloat[The real part of arccotangent with a branch cut at ${[-\infty\iunit, -\iunit]}$, ${[\iunit,\infty\iunit]}$.]{%
        \includegraphics[width=0.45\textwidth]{acotCut1.png}
        \label{fig:acot-cut1}%
    }
    \hspace{0.5cm}
    \subfloat[The real part of arccotangent with a branch cut at ${[-\iunit, \iunit]}$.]{%
        \includegraphics[width=0.45\textwidth]{acotCut2.png}
        \label{fig:acot-cut2}%
    }
    \caption{Two plots for the real part of the arccotangent function with a branch cut at $[-\infty\iunit, -\iunit]$, $[\iunit,\infty\iunit]$ in figure~\protect\subref{fig:acot-cut1} and at $[-\iunit, \iunit]$ in figure~\protect\subref{fig:acot-cut2}, respectively. (Plotted with \Maple{} 2016)}
    \label{fig:acot-cut-compare}
\end{figure}

@@

This article presents a new approach for automatic \gls*{p2c} and vice versa conversions. Translations from presentational to computational (computational to presentational) systems are called forward (backward) translations. A forward translation is denoted with an arrow with the target system language above the arrow. For example,
\begin{equation*}
t \overset{\langMaple}{\mapsto} c,
\end{equation*}
where $t$ is an expression in the \LaTeX{} language and $c$ is an element of the \Maple{} language $\langMaple$. As we will see later in this article, we need to compare mathematical concepts between systems. This is impossible from a mathematical point of view. Consider the irrational mathematical constant $e$, known as Euler's number. The theoretical construct for this symbol cannot be mathematically equivalent to the value \verb|exp(0)| in \Maple, caused by computational and implementational limitations. Instead of using the term \textit{equivalent}, we introduce a \textit{appropriate} and \textit{inappropriate} translations. We call a translation such as
\begin{equation}\label{eq:cos-def}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|cos(z)|
\end{equation}
as \textit{appropriate}, while a translation such as
\begin{equation}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|sin(z)|
\end{equation}
is called \textit{inappropriate}. Note that it is not always as easy as in this example to decide if a translation is appropriate or not. Therefore, this article also presents several validation techniques to automatically verify if a translation is appropriate or inappropriate. In addition to this terminology, we introduce \textit{direct translations}. Later in the paper, we will explain that a translation from one specific mathematical object to its \textit{appropriate} counterpart in the other system is not always possible. We call a translation to the \textit{appropriate} counterpart \textit{direct}. For example, the translation~(\ref{eq:cos-def}) is \textit{direct}, while a translation to the definition of the cosine function
\begin{equation*}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|(exp(I*z)+exp(-I*z))/2|
\end{equation*}
is not a \textit{direct} translation.

>>>>

Even with embedded semantic information, a translation between systems can be difficult. A typical example of complex problems occur for multivalued functions~\parencite{AISC:MultivaluedFunctions}. A \gls*{cas} usually defines \textit{branch cuts} to compute principal values of multivalued functions~\parencite{Maple:Cuts}, which makes the implementation of a theoretically continued function to a discontinuous presentation of it. In general, positioning branch cuts follows conventions, but can be positioned arbitrarily in many cases. Communicating and explaining the decision of defined branch cuts is a critical issue for \gls*{cas} and can vary between various systems~\parencite{Branches:acot}. Figure~\ref{fig:acot-cut-compare} illustrates two examples of different branch cut positioning for the inverse trigonometric arccotangent function. While \Maple{} defines the branch cut at ${[-\iunit\infty, -\iunit]}$, ${[\iunit,\iunit\infty]}$ (Figure~\ref{fig:acot-cut1}), \Mathematica{} defines the branch cut at ${[-\iunit, \iunit]}$ (figure~\ref{fig:acot-cut2}).

\begin{figure}[ht]
    \centering
    \subfloat[The real part of arccotangent with a branch cut at ${[-\iunit\infty, -\iunit]}$, ${[\iunit,\iunit\infty]}$.]{%
        \includegraphics[width=0.45\textwidth]{acotCut1.png}
        \label{fig:acot-cut1}%
    }
    \hspace{0.5cm}
    \subfloat[The real part of arccotangent with a branch cut at ${[-\iunit, \iunit]}$.]{%
        \includegraphics[width=0.45\textwidth]{acotCut2.png}
        \label{fig:acot-cut2}%
    }
    \caption{Two plots for the real part of the arccotangent function with a branch cut at $[-\iunit\infty, -\iunit]$, $[\iunit,\iunit\infty]$ in figure~\protect\subref{fig:acot-cut1} and at $[-\iunit, \iunit]$ in figure~\protect\subref{fig:acot-cut2}, respectively. (Plotted with \Maple{} 2016)}
    \label{fig:acot-cut-compare}
\end{figure}

@@

This article presents a new approach for automatic \gls*{p2c} and vice versa conversions. Translations from presentational to computational (computational to presentational) systems are called forward (backward) translations. A forward translation is denoted with an arrow with the target system language above the arrow. For example,
\begin{equation*}
t \overset{\langMaple}{\mapsto} c,
\end{equation*}
where $t$ is an expression in the \LaTeX{} language and $c$ is an element of the \Maple{} language $\langMaple$. As we will see later in this article, we need to compare mathematical concepts between systems. This is impossible from a mathematical point of view. Consider the irrational mathematical constant $e$, known as Euler's number. The theoretical construct for this symbol cannot be mathematically equivalent to the value \verb|exp(0)| in \Maple, caused by computational and implementational limitations. Instead of using the term \textit{equivalent}, we introduce a \textit{appropriate} and \textit{inappropriate} translations. We call a translation such as
\begin{equation}\label{eq:cos-def}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|cos(z)|
\end{equation}
as \textit{appropriate}, while a translation such as
\begin{equation}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|sin(z)|
\end{equation}
as \textit{inappropriate}. Note that it is not always as easy as in this example to decide if a translation is appropriate or not. Therefore, this article also presents several validation techniques to automatically verify if a translation is appropriate or inappropriate. In addition to this terminology, we introduce \textit{direct translations}. Later in the paper, we will explain that a translation from one specific mathematical object to its \textit{appropriate} counterpart in the other system is not always possible. We call a translation to the \textit{appropriate} counterpart \textit{direct}. For example, the translation~(\ref{eq:cos-def}) is \textit{direct}, while a translation to the definition of the cosine function
\begin{equation*}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|(exp(I*z)+exp(-I*z))/2|
\end{equation*}
is not a \textit{direct} translation.

### 2018-07-30 chapters/02-background/related-work.tex
Implemented all suggestions by Howard

\section{Related Work}\label{sec:related-work}
Since \LaTeX{} became the de facto standard for writing papers in mathematics, most of the \gls*{cas} provide simple functions to import and export mathematical \LaTeX{} expressions%
\footnote{The selected \gls*{cas} Maple, Mathematica, Matlab, and SageMath provide import and/or export functions for \LaTeX:\\
Maple, \url{http://www.maplesoft.com/support/help/Maple/view.aspx?path=latex} seen 06/2017;\\
Mathematica, \url{https://reference.wolfram.com/language/tutorial/GeneratingAndImportingTeX.html} seen 06/2017;\\
Matlab, \url{https://www.mathworks.com/help/symbolic/latex.html} seen 06/2017;\\
SageMath, \url{http://doc.sagemath.org/html/en/tutorial/latex.html} seen 06/2017}. %
Those tools have two essential problems. They are only able to import simple mathematical expressions, where the semantics are unique. For example, the internal \LaTeX{} macro \verb|\frac| always indicates a fraction. However, for more complex expressions, e.g., the Jacobi polynomial in \cref{tab:JacobiP-usecase}, the import functions fail. The second problem appears in the export tools. Mathematical expressions in \gls*{cas} are fully semantic. Otherwise the \gls*{cas} wouldn't be able to compute or evaluate the expressions. During the export process, the semantic information gets lost, because generic \LaTeX{} is not able to carry semantic information. In consequence of these two problems, an exported expression cannot be imported to the same system again in most cases (except those simple expressions described above). Our tool should solve these problems and provide round-trip translations between \LaTeX{} and \gls*{cas}.

The semantics must be well known before an expression can be translated. There are two main approaches to solve that problem: (1) someone could specify the semantic information during the writing process (pre-defined semantics) or (2) the translator can find out the right semantic information in general mathematical expressions before it translates the expression. So-called \textit{interactive documents}\footnote{There is no adequate definition what interactive documents are. However, this name is widely used to describe electronic document formats that allow interactivity to change the content in real time.}, Such as the \gls*{cdf}~\parencite{CDF:Wolfram} by Wolfram Research or the \textit{worksheets} by \Maple{}, try to solve this problem with the approach (2) and allow to embed semantic information into the input. Those complex document formats require specialized tools to show and work with the documents (Wolfram CDF Player, or \Maple{} for the \textit{worksheets}). The \JOBAD{} architecture~\parencite{JOBAD:orig} is able to create web-based interactive documents and uses \gls*{omdoc}~\parencite{OMDoc} to carry semantics. The documents can be viewed and edited in the browser. Those \JOBAD{-documents} also allow to perform computations during \gls*{cas}. This gives the opportunity to calculate, compute and change mathematical expressions directly in the document. The translation performs in the background, invisible to the user. Similar to the \JOBAD{} architecture other interactive web documents exist, such as \textit{MathDox}~\parencite{MathDox} and \textit{The Planetary System}~\parencite{Planetary}. All of them demonstrate the potential of the education system.

Another approach tries to avoid translation problems by allow computations directly via the \LaTeX{} compiler, e.g., \textit{LaTeXCalc}~\parencite{LatexCalc}. Those packages are limited to the possibilities of the compiler and therefore not as powerful as \gls*{cas}. A workaround for this case is \textit{sagetex}~\parencite{Sagetex}, which is a \LaTeX{} package interface for the open source \gls*{cas} \textit{sage}\footnote{An abbreviation for \textit{SageMath}.}. This package allows \textit{sage} commands in \TeX{-}files and uses \textit{sage} in the background to compute the commands. In this scenario, a writer still needs to manually translate expressions to the syntax of \textit{sage}.

There exist two approaches that allow embedding semantic information within \LaTeX{} expressions by using custom macros. Namely, \sTeX{}~\parencite{sTeX} developed by Kohlhase and the \Macro s developed by Miller~\parencite{DLMF:Macros}. This paper shows that it is possible to develop a context-free translation tool using the semantic macros introduced by these two projects. \sTeX{} aims to embed semantic information for general mathematics with a comprehensive set of macros. The macro set developed by Miller introduces new macros for special functions, orthogonal polynomials, and mathematical constants. Each of these macros ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf} or \gls*{drmf}. Therefore, we call these semantic macros \Macro s. These semantic macros are internally used in the \gls*{dlmf} and the \gls*{drmf}. Because of the linked definitions to the \gls*{dlmf} and the special macros for functions and mathematical constants this paper using the \Macro s for performing translations to \gls*{cas} instead of using \sTeX.

>>>>

\section{Related Work}\label{sec:related-work}
Since \LaTeX{} became the de facto standard for writing papers in mathematics, most of the \gls*{cas} provide simple functions to import and export mathematical \LaTeX{} expressions%
\footnote{The selected \gls*{cas} Maple, Mathematica, Matlab, and SageMath provide import and/or export functions for \LaTeX:\\
Maple, \url{http://www.maplesoft.com/support/help/Maple/view.aspx?path=latex} seen 06/2017;\\
Mathematica, \url{https://reference.wolfram.com/language/tutorial/GeneratingAndImportingTeX.html} seen 06/2017;\\
Matlab, \url{https://www.mathworks.com/help/symbolic/latex.html} seen 06/2017;\\
SageMath, \url{http://doc.sagemath.org/html/en/tutorial/latex.html} seen 06/2017}. %
Those tools have two essential problems. They are only able to import simple mathematical expressions, where the semantics are unique. For example, the internal \LaTeX{} macro \verb|\frac| always indicates a fraction. However, for more complex expressions, e.g., the Jacobi polynomial in \Cref{tab:JacobiP-usecase}, the import functions fail. The second problem appears in the export tools. Mathematical expressions in \gls*{cas} are fully semantic. Otherwise the \gls*{cas} wouldn't be able to compute or evaluate the expressions. During the export process, the semantic information gets lost, because generic \LaTeX{} is not able to carry sufficient semantic information. In consequence of these two problems, an exported expression cannot be imported to the same system again in most cases (except for simple expressions such as those described above). Our tool should solve these problems and provide round-trip translations between \LaTeX{} and \gls*{cas}.

The semantics must be well known before an expression can be translated. There are two main approaches to solve that problem: (1) someone could specify the semantic information during the writing process (pre-defined semantics); and (2) the translator can find out the right semantic information in general mathematical expressions before it translates the expression. So-called \textit{interactive documents}\footnote{There is no adequate definition what interactive documents are. However, this name is widely used to describe electronic document formats that allow for interactivity to change the content in real time.}, such as the \gls*{cdf}~\parencite{CDF:Wolfram} by Wolfram Research, or \textit{worksheets} by \Maple{}, try to solve this problem with the approach (2) and allow one to embed semantic information into the input. Those complex document formats require specialized tools to show and work with the documents (Wolfram CDF Player, or \Maple{} for the \textit{worksheets}). The \JOBAD{} architecture~\parencite{JOBAD:orig} is able to create web-based interactive documents and uses \gls*{omdoc}~\parencite{OMDoc} to carry semantics. The documents can be viewed and edited in the browser. Those \JOBAD{-documents} also allow one to perform computations via \gls*{cas}. This gives one the opportunity to calculate, compute and change mathematical expressions directly in the document. The translation performs in the background, invisible to the user. Similar to the \JOBAD{} architecture, other interactive web documents exist, such as \textit{MathDox}~\parencite{MathDox} and \textit{The Planetary System}~\parencite{Planetary}. All of these demonstrate the potential of the educational system.

Another approach tries to avoid translation problems by allowing computations directly via the \LaTeX{} compiler, e.g., \textit{LaTeXCalc}~\parencite{LatexCalc}. Those packages are limited to the abilities of the compiler and therefore are not as powerful as \gls*{cas}. A workaround for this case is \textit{sagetex}~\parencite{Sagetex}, which is a \LaTeX{} package interface for the open source \gls*{cas} \textit{sage}\footnote{An abbreviation for \textit{SageMath}.}. This package allows \textit{sage} commands in \TeX{-}files and uses \textit{sage} in the background to compute the commands. In this scenario, a writer still needs to manually translate expressions to the syntax of \textit{sage}.

There exist two approaches for marking up mathematical \TeX/\LaTeX{} documents semantically with \TeX{} macros. Namely, \sTeX{}~\parencite{sTeX} developed by Kohlhase and the \Macro s developed by Miller~\parencite{DLMF:Macros}. This paper shows that it is possible to develop a context-free translation tool using the semantic macros introduced by these two projects. The goal of \sTeX{} was to markup the functional structure of mathematical documents so that they can be exported to the \gls*{omdoc} format. The macro set developed by Miller introduces new macros for special functions, orthogonal polynomials, and mathematical constants. Each of these macros ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf} or \gls*{drmf}. Therefore, we call these semantic macros \Macro s. These semantic macros are internally used in the \gls*{dlmf} and the \gls*{drmf}. We gave the \Macro{} set the favor for developing the translation engine because it provides \gls*{dlmf} definitions for a comprehensive amount of functions. In contrast, \sTeX{} does not focus on the semantics of functions, is too complex to use, and defines diverse macros for symbols and concepts that \gls*{cas} usually does not support.

%Because of the linked definitions to the \gls*{dlmf} and the special macros for functions and mathematical constants this paper using the \Macro s for performing translations to \gls*{cas} instead of using \sTeX.

### 2018-07-30 chapters/03-translator/backward-trans.tex
Implemented all suggestions by Howard

One can access the internal data structure of expressions via the \texttt{ToInert} command. The returned \inertF{} is a nested list representation as a tree equivalent\footnote{A tree equivalent of a \gls*{dag} splitting nodes with multiple parents into multiple nodes so that each node has only one parent node.} of the internal \gls*{dag} for the given expression. Some of the important types for the nodes are specified in table~\ref{tab:maple-types}.

@@

\subsection{Automatic Changes of Inputs in Maple}\label{subsec:maple-probs}
\Maple{} evaluates inputs automatically and changes the input into an internal representation. This internal representation might look a bit different to the input. One example has already been given with figure~\ref{fig:internal-maple-dag}, where each summand of a sum is stored with a factor. Here is a list of all internal changes that occur for inputs.

@@

The translation process then follows the same principle as for the forward translations. Since the syntax tree of \Maple{} is an expression tree, we do not need to implement special reordering or grouping algorithms to perform backward translations. Translations for functions are also realized via patterns and placeholders. Figure~\ref{fig:backward-trans} illustrates the backward translation process for the Jacobi polynomial example from table~\ref{tab:JacobiP-usecase}.

\begin{figure}[!htp]
	\centering
	\includegraphics[clip, trim=0.1cm 0.1cm 0.1cm 0.1cm, scale=0.7]{MapleTranslation.pdf}
	\caption{A scheme of the backward translation process from \Maple{} for the Jacobi polynomial $P_n^{(\alpha , \beta)}(\cos(a\Theta))$. The input string is converted by the \Maple{} kernel into the nested list representation. This list is translated by subtranslators (blue and red arrows). A function translation (bold blue arrows) is again realized by translation patterns to define the position of the arguments (red arrows).}
	\label{fig:backward-trans}
\end{figure}

>>>>

One can access the internal data structure of expressions via the \texttt{ToInert} command. The returned \inertF{} is a nested list representation as a tree equivalent\footnote{A tree equivalent of a \gls*{dag} splitting nodes with multiple parents into multiple nodes so that each node has only one parent node.} of the internal \gls*{dag} for the given expression. Some of the important types for the nodes are specified in \Cref{tab:maple-types}.

@@

\subsection{Automatic Changes of Inputs in Maple}\label{subsec:maple-probs}
\Maple{} evaluates inputs automatically and changes the input into an internal representation. This internal representation might look a bit different to the input. One example has already been given with \Cref{fig:internal-maple-dag}, where each summand of a sum is stored with a factor. Here is a list of all internal changes that occur for inputs.

@@

The translation process then follows the same principle as for the forward translations. Since the syntax tree of \Maple{} is an expression tree, we do not need to implement special reordering or grouping algorithms to perform backward translations. Translations for functions are also realized via patterns and placeholders. Figure~\ref{fig:backward-trans} illustrates the backward translation process for the Jacobi polynomial example from \Cref{tab:JacobiP-usecase}.

\begin{figure}[t!]
	\centering
	\includegraphics[clip, trim=0.1cm 0.1cm 0.1cm 0.1cm, scale=0.7]{MapleTranslation.pdf}
	\caption{A scheme of the backward translation process from \Maple{} for the Jacobi polynomial $P_n^{(\alpha , \beta)}(\cos(a\Theta))$. The input string is converted by the \Maple{} kernel into the nested list representation. This list is translated by subtranslators (blue and red arrows). A function translation (bold blue arrows) is again realized by translation patterns to define the position of the arguments (red arrows).}
	\label{fig:backward-trans}
\end{figure}
